Finding Maximum $k$-Cliques Faster using Lazy Global Domination

### 2016-04-06 paper/paper.tex
Tweaks

\begin{abstract}
    A clique in a graph is a set of vertices, each of which is adjacent to every other vertex in
    this set. A $k$-clique relaxes this requirement, requiring vertices to be within a distance $k$
    of each other, rather than directly adjacent. In theory, a maximum clique algorithm can easily
    be adapted to solve the maximum $k$-clique problem, although large sparse $k$-clique graphs
    reduce to large dense clique graphs, which can be computationally challenging. We adapt a state of
    the art maximum clique algorithm to show that this reduction is in fact useful in practice, and
    introduce a lazy global domination rule which sometimes vastly reduces the search space. We
    include experimental results for a range of real-world and benchmark graphs, and a detailed look
    at random graphs.
\end{abstract}

@@

Here we give experimental results on a range of standard benchmarks, and on real-world and random
graphs. Where timing results are reported, the experiments were run on a machine with dual Intel
E5-2687W v3 processors; single-threaded runtimes are given, except in \cref{table:parallel}, where
20 threads were used (our machine does not have hyper-threading enabled). The time taken to read in
the graph from a file is excluded, but preprocessing time (including the construction of $G^k$ and
the bitset encoding) is included. We use the term \emph{nodes} to refer to the number of recursive
calls made by the branch-and-bound part of the algorithm.

@@

\begin{table}
    \tiny
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-erdos}
    \bottomrule
\end{tabular}
\caption{Experimental results for Erd\H{o}s collaboration graphs. For each
graph, we consider $k$ equal to 2, 3 and 4. In each case we show the density of
$G^k$, the size of a maximum $k$-clique, and then for both the unmodified
algorithm and the algorithm with our lazy global domination step, the number of
nodes required, and the runtime in seconds.}\label{table:erdos}
\end{table}

@@

\begin{table}
    \tiny
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-dimacs}
    \bottomrule
\end{tabular}
\caption{Experimental results for DIMACS clique graphs with diameter greater
    than two. For each graph, we consider $k$ equal to 2, 3 and 4. In each case
    we show the density of $G^k$, the size of a maximum $k$-clique, and then
    for both the unmodified algorithm and the algorithm with our lazy global
domination step, the number of nodes required, and the runtime in seconds.}\label{table:clique}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{4pt}
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-dimacs10walshaw}
    \bottomrule
\end{tabular}
\caption{Experimental results for the smaller DIMACS partitioning graphs. For
    each graph, we consider $k$ equal to 2, 3 and 4. In each case we show the
    density of $G^k$, the size of a maximum $k$-clique, and then for both the
    unmodified algorithm and the algorithm with our lazy global domination
step, the number of nodes required, and the runtime in seconds.}\label{table:partitioning}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{4pt}
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-dimacs10cluster}
    \bottomrule
\end{tabular}
\caption{Experimental results for the smaller DIMACS clustering graphs. For
    each graph, we consider $k$ equal to 2, 3 and 4. In each case we show the
    density of $G^k$, the size of a maximum $k$-clique, and then for both the
    unmodified algorithm and the algorithm with our lazy global domination
step, the number of nodes required, and the runtime in seconds.}\label{table:clustering}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{4pt}
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Sequential} & \multicolumn{2}{c}{Parallel} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-parallel}
    \bottomrule
\end{tabular}
\caption{Experimental results for multi-threaded search on harder instances, using 20 threads. For
    each graph, we consider $k$ equal to 2, 3 and 4. In each case we show the
    density of $G^k$, the size of a maximum $k$-clique, and then for both the
    sequential algorithm and the parallel algorithm (with lazy global domination in both cases),
    the number of nodes required, and the runtime in seconds.}\label{table:parallel}
\end{table}

>>>>

\begin{abstract}
    A clique in a graph is a set of vertices, each of which is adjacent to every other vertex in
    this set. A $k$-clique relaxes this requirement, requiring vertices to be within a distance $k$
    of each other, rather than directly adjacent. In theory, a maximum clique algorithm can easily
    be adapted to solve the maximum $k$-clique problem, although large sparse $k$-clique graphs
    reduce to large dense clique graphs, which can be computationally challenging. We adapt a state of
    the art maximum clique algorithm to show that this reduction is in fact useful in practice, and
    introduce a lazy global domination rule which sometimes vastly reduces the search space. We
    include experimental results for a range of real-world and benchmark graphs, and a detailed look
    at random graphs. We also use thread-parallel search to solve some harder instances.
\end{abstract}

@@

Here we give experimental results on a range of standard benchmarks, and on real-world and random
graphs. Where timing results are reported, the experiments were run on a machine with dual Intel
E5-2687W v3 processors and 256GBytes RAM; single-threaded runtimes are given, except in
\cref{table:parallel}, where 20 threads were used (our machine does not have hyper-threading
enabled). Our software was implemented in C++, using C++11 native threads, and was compiled using
GCC 5.1.0. The time taken to read in the graph from a file is excluded, but preprocessing time
(including the construction of $G^k$ and the bitset encoding) is included. We use the term
\emph{nodes} to refer to the number of recursive calls made by the branch-and-bound part of the
algorithm.

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{6pt} % carefully tuned!
    \centering
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-erdos}
    \bottomrule
\end{tabular}
\caption{Experimental results for Erd\H{o}s collaboration graphs. For each
graph, we consider $k$ equal to 2, 3 and 4. In each case we show the density of
$G^k$, the size of a maximum $k$-clique, and then for both the unmodified
algorithm and the algorithm with our lazy global domination step, the number of
nodes required, and the runtime in seconds.}\label{table:erdos}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{6pt} % carefully tuned!
    \centering
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-dimacs}
    \bottomrule
\end{tabular}
\caption{Experimental results for DIMACS clique graphs with diameter greater
    than two. For each graph, we consider $k$ equal to 2, 3 and 4. In each case
    we show the density of $G^k$, the size of a maximum $k$-clique, and then
    for both the unmodified algorithm and the algorithm with our lazy global
domination step, the number of nodes required, and the runtime in seconds.}\label{table:clique}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{4pt} % carefully tuned!
    \centering
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-dimacs10walshaw}
    \bottomrule
\end{tabular}
\caption{Experimental results for the smaller DIMACS partitioning graphs. For
    each graph, we consider $k$ equal to 2, 3 and 4. In each case we show the
    density of $G^k$, the size of a maximum $k$-clique, and then for both the
    unmodified algorithm and the algorithm with our lazy global domination
step, the number of nodes required, and the runtime in seconds.}\label{table:partitioning}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{5pt} % carefully tuned!
    \centering
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Unmodified} & \multicolumn{2}{c}{With Domination} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-dimacs10cluster}
    \bottomrule
\end{tabular}
\caption{Experimental results for the smaller DIMACS clustering graphs. For
    each graph, we consider $k$ equal to 2, 3 and 4. In each case we show the
    density of $G^k$, the size of a maximum $k$-clique, and then for both the
    unmodified algorithm and the algorithm with our lazy global domination
step, the number of nodes required, and the runtime in seconds.}\label{table:clustering}
\end{table}

@@

\begin{table}
    \tiny\setlength{\tabcolsep}{4.6pt} % carefully tuned!
    \centering
    \begin{tabular}{l c rr rr rr}
        \toprule
        & & & & \multicolumn{2}{c}{Sequential} & \multicolumn{2}{c}{Parallel} \\
    \cmidrule(lr){5-6}
    \cmidrule(lr){7-8}
    Instance & \multicolumn{1}{c}{$k$} & \multicolumn{1}{c}{$D$} & \multicolumn{1}{c}{$\tilde{\omega}_k$} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} &
    \multicolumn{1}{c}{Nodes} & \multicolumn{1}{c}{Time} \\
    \midrule
    \inputhaxx{gen-table-parallel}
    \bottomrule
\end{tabular}
\caption{Experimental results for multi-threaded search on harder instances, using 20 threads. For
    each graph, we consider $k$ equal to 2, 3 and 4. In each case we show the
    density of $G^k$, the size of a maximum $k$-clique, and then for both the
    sequential algorithm and the parallel algorithm (with lazy global domination in both cases),
    the number of nodes required, and the runtime in seconds.}\label{table:parallel}
\end{table}
